An Ordered Logic for Category Theory

### 2022-09-07 paper/paper.tex
talk about extensionality

>>>>

\subsection{Extensionality Axioms}

Finally, we discuss some non-theorems that would be desirable to add
as axioms to the theory of \ohol{}.
%
Chief among these are principles about the hom type $\punitinXfromYtoZ
{\cat C} \alpha \beta$ when $\cat C$ is a connective.
%
Since the hom type is defined as a kind of unit type, rather than
defined inductively, we are not guaranteed that the hom type for the
product category $\cat C \times \cat D$ or the presheaf categories are
what we expect.
%
For the product type, we expect there should be an isomorphism
$\punitinXfromYtoZ {\cat C \times \cat D} \alpha \beta \cong
(\punitinXfromYtoZ {\cat C} {\pi_1 \alpha} {\pi_1 \beta}) \times
(\punitinXfromYtoZ {\cat D} {\pi_2 \alpha} {\pi_2 \beta})$ saying an
arrow in a product category is a pair of arrows.
%
Similarly, the hom type in the presheaf category should be given by
the end formula $\punitinXfromYtoZ {\negPresheaf {\cat C}} \alpha
\beta \cong \homlallXYtoZ {\gamma}{\negPresheafApp \alpha
  \gamma}{\negPresheafApp \beta \gamma}$. This would allow us to
relate the Yoneda Lemma \ref{lem:yoneda} to the morphisms of
presheaves, and prove that the Yoneda embedding is fully faithful.
%
We call these extensionality principles because they are analogous to
principles such as function extensionality but for the directed hom
type rather than equality.

The presheaf extensionality principle is independent from the existing
axioms of \ohol{}: it is true in the ordinary category theoretic
model, but it is false in the set theoretic model. That is, in the set
theoretic model, the presheaf category is simply the powerset, the hom
type is \emph{equality} and the function profunctor is interpreted as
a universally quantified \emph{implication}, but equality in the
powerset should be \emph{bi-implication}, so the theorem is false. We
expect the product extensionality is independent as well, but a
counter-model would likely be quite exotic as we expect it to be
provable in a system with functor categories, in analogy with the
proof for the identity type in MLTT.
%
Note that these axioms can easily be added as assumptions in the
dependently typed context, though this will break the constructive
character of the system. We expect techniques from type theory might
be applied to incorporate these as computationally meaningful
principles rather than assumed as axioms \cite{obs-tt,xtt,cubical-type-theory}.
